---
title: "An investigation of the long term ecological and soil carbon storage responses of the Upper Axe valley to landscape scale anthropogenic impacts"
author: "Julian Payne"
date: "2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


### MSc Land and Ecological Restoration - Research Project
# Consideration of soil horizon classification
## detailed analysis of combined gley soils (with gleys in wetlands removed)
### with reference to sample depth below the surface (depth_cm)

```{r}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```

### Load packages
```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(readxl)
library(cowplot)
library(vegan)
library(lme4)
library(lmerTest)
```

## Load the data for the project
```{r}
filepath<-here("data","Soil_Cores.xlsx")
soil_cores<-read_excel(filepath,
                       sheet = "Soildata",
                       range = "A1:AB155")
```
### Set soil wetness layer order
```{r}
WetnessOrder<-c("Dry", "Damp", "Wet")
```

### Clean up the data
Notice some missing values in the loi column, identified by 0 in test_for_loi

```{r}
soilloi_clean<-soil_cores %>%
  filter( horizon_type == "Floodplain1" |horizon_type == "Floodplain2" | horizon_type == "Gley" | horizon_type == "Wetland" | horizon_type == "RiverBed") %>%  
  select(-loi) %>%
  drop_na()
```

# considering Gley soils in detail

```{r}
soilloi_clean%>%
  filter(River=="Axe")%>%
  filter(horizon_type=="Gley")%>%
  ggplot(aes(x=depth_cm, y=loi_calc)) +
  geom_point(aes(size=wetdrydamp), alpha=0.5, colour="Blue")+
    labs(x = "Depth (cm) below surface",
       y = "Total Carbon LOI",
       title = "Variation of Gley soil LOI with Depth for the Axe Valley",
          size = "Soil wetness") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))+
  theme(legend.position = c(0.83,0.8), legend.title = element_text(size =12), legend.text = element_text(size=10))
```


```{r}
soilloi_clean%>%
  filter(River=="Axe")%>%
  filter(horizon_type=="Gley")%>%
  ggplot(aes(x=depth_cm, y=loi_calc, colour = factor( wetdrydamp, ordered(WetnessOrder)))) +
  geom_point(aes(size=core_type), alpha=0.5)+
    labs(x = "Depth (cm) below surface",
       y = "Total Carbon LOI",
       title = "Variation of Gley soil LOI with Depth for the Axe Valley",
       size = "Core location",
       colour = "Soil wetness") +
  theme_cowplot() +
  theme(title = element_text(size=10.5))+
  theme(legend.position = c(0.8,0.75), legend.title = element_text(size =10.5), legend.text = element_text(size=10))
```


```{r}
soilloi_clean%>%
  filter(River=="Axe")%>%
  filter(horizon_type=="Gley")%>%
  ggplot(aes(x=depth_cm, y=loi_calc, colour = core_type )) +
  geom_point(aes(size=factor( wetdrydamp, ordered(WetnessOrder))), alpha=0.5)+
    labs(x = "Depth (cm) below surface",
       y = "Total Carbon LOI",
       title = "Variation of Gley soil LOI with Depth for the Axe Valley",
       size = "Soil wetness",
       colour = "Core location") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))+
  theme(legend.position = c(0.8,0.75), legend.title = element_text(size =10.5), legend.text = element_text(size=10))
```

```{r}
soilloi_clean%>%
  filter(River=="Axe")%>%
  filter(horizon_type=="Gley")%>%
  ggplot(aes(x=core_type, y=loi_calc)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour ="#f03b00" ,alpha=0.5) +
    labs(x = "Core location",
       y = "Carbon LOI",
       title = "Variation of Gley LOI with core location") +
  theme_cowplot() 
```
















