---
title: "An investigation of the long term ecological and soil carbon storage responses of the Upper Axe valley to landscape scale anthropogenic impacts"
author: "Julian Payne"
date: "2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### MSc Land and Ecological Restoration - Research Project
# Variation of LOI with soil horizon, and by river and soil wetness
### Wetlands split into BuriedWetlands (including RiverBed) located underneath grey clays and Wetlands for contemporary overlying wetland soils and associated grey clay

```{r}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```

### Load packages
```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(readxl)
library(cowplot)
library(vegan)
library(lme4)
library(lmerTest)
```

### Load the data for the project
```{r}
filepath<-here("data","Soil_Cores.xlsx")
soil_cores<-read_excel(filepath,
                       sheet = "SoildataV3",
                       range = "A1:AB155")
```

## Set soil horizon layer order
```{r}
HorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Wetland", "BuriedWetland")
PaleoHorizonOrder<-c("Floodplain1", "Gley", "Wetland", "Buried Wetland")
FldPlnHorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Buried Wetland")
WetnessOrder<-c("Dry", "Damp", "Wet")
```

### Clean up the data
Notice we have some missing values in the loi column, identified by 0 in test_for_loi

```{r}
soilloi_clean<-soil_cores %>%
  filter(horizon_type == "Floodplain2" | horizon_type == "Floodplain1" | horizon_type == "Gley" | horizon_type == "Wetland" | horizon_type == "BuriedWetland") %>% 
  select(-loi) %>%
  drop_na()
```

## split floodplain and paleochannel data
```{r}
paleochannel<-soilloi_clean %>%
  filter(core_type=="Paleochannel")
floodplain<-soilloi_clean %>%
  filter(core_type=="Floodplain")
```

## Plot the data

```{r}
soilloi_clean%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=loi_calc, ymax =0.35, ymin = 0, colour =River)) +
  geom_boxplot()+
#  geom_jitter(width=0.1, colour ="#f03b00" ,alpha=0.5) +
  geom_point(position = position_jitterdodge())+
    labs(x = "Soil Horizon",
       y = "Total Carbon LOI",
       title = "Variation of LOI by river and soil horizon") +
  theme_cowplot() +
  theme(legend.position = c(0.05,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10))
```

```{r}
paleochannel%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=loi_calc, ymax = 0.35, ymin = 0, colour =River)) +
  geom_boxplot()+
  #  geom_jitter(width=0.1, colour ="#f03b00" ,alpha=0.5) +
  geom_point(position = position_jitterdodge())+
    labs(x = "Soil Horizon",
       y = "Total Carbon LOI",
       title = "Total carbon for Paleochannel core soil samples") +
  theme_cowplot() +
  theme(legend.position = c(0.05,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10))

```


```{r}
floodplain%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=loi_calc, ymax = 0.35, ymin = 0, colour =River)) +
  geom_boxplot()+
  #  geom_jitter(width=0.1, colour ="#f03b00" ,alpha=0.5) +
  geom_point(position = position_jitterdodge())+
    labs(x = "Soil Horizon",
       y = "Total Carbon LOI",
       title = "Total carbon for Floodplain core soil samples") +
  theme_cowplot() +
  theme(legend.position = c(0.05,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10))
```


## Test if soil horizon data sets are significantly different for each river

```{r}
AxeFdpn1Loi<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull() 

SynFdpn1Loi<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull()  

AxeGleyLoi<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull() 

SynGleyLoi<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull() 

AxeWetlandLoi<-soilloi_clean %>%
  filter(horizon_type=="Wetland") %>%
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull()

SynWetlandLoi<-soilloi_clean %>%
  filter(horizon_type=="Wetland") %>%
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull()

AxeBuriedWetlandLoi<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull()

SynBuriedWetlandLoi<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull()

t.test(AxeFdpn1Loi,SynFdpn1Loi)
t.test(AxeGleyLoi,SynGleyLoi)
t.test(AxeWetlandLoi,SynWetlandLoi)
t.test(AxeBuriedWetlandLoi, SynBuriedWetlandLoi)

```
### Consider the influence of soil wetness

## Plot Wetland data

```{r}
soilloi_clean%>%
  filter(horizon_type=="Wetland")%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=loi_calc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour ="#f03b00" ,alpha=0.5) +
    labs(x = "Soil wetness",
       y = "Total Carbon LOI",
       title = "Total Carbon of Wetland soils by soil wetness") +
  theme_cowplot() 
```

```{r}
soilloi_clean%>%
  filter(horizon_type=="Wetland")%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=loi_calc, ymax =0.35, ymin = 0, colour =River)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
    labs(x = "Soil wetness",
       y = "Carbon LOI",
       title = "Total Carbon of Wetland soils by soil wetness and river") +
  theme_cowplot() +
  theme(legend.position = c(0.05,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10))
```

## Plot BuriedWetland data

```{r}
soilloi_clean%>%
  filter(horizon_type=="BuriedWetland")%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=loi_calc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour ="#f03b00" ,alpha=0.5) +
    labs(x = "Soil wetness",
       y = "Total Carbon LOI",
       title = "Total Carbon of Buried Wetland soils by soil wetness") +
  theme_cowplot() 
```

```{r}
soilloi_clean%>%
  filter(horizon_type=="BuriedWetland")%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=loi_calc, ymax =0.35, ymin = 0, colour =River)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
    labs(x = "Soil wetness",
       y = "Total Carbon LOI",
       title = "Total Carbon of Buried Wetland soils by soil wetness and river") +
  theme_cowplot() +
  theme(legend.position = c(0.05,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10))
```

## Test if Buried Wetland soil of different wettness are significantly different

```{r}
BuriedWetDampLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Damp") %>% 
  filter(horizon_type=="BuriedWetland") %>%
  select(loi_calc) %>%
  pull() 

BuriedWetDryLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Dry") %>% 
  filter(horizon_type=="BuriedWetland") %>%
  select(loi_calc) %>%
  pull() 

BuriedWetWetLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Wet") %>% 
  filter(horizon_type=="BuriedWetland") %>%
  select(loi_calc) %>%
  pull() 


t.test(BuriedWetDryLoi,BuriedWetDampLoi)
t.test(BuriedWetDryLoi,BuriedWetWetLoi)
t.test(BuriedWetDampLoi,BuriedWetWetLoi)

```

## Test if soil of different wettness are significantly different for each river


```{r}
AxeDampLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Damp") %>% 
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull() 

SynDampLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Damp") %>% 
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull()  

AxeDryLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Dry") %>% 
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull() 

SynDryLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Dry") %>% 
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull()  

AxeWetLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Wet") %>% 
  filter(River=="Axe") %>%
  select(loi_calc) %>%
  pull() 

SynWetLoi<-soilloi_clean %>%
  filter(wetdrydamp=="Wet") %>% 
  filter(River=="Synderford") %>%
  select(loi_calc) %>%
  pull()  


t.test(AxeDryLoi,SynDryLoi)
t.test(AxeDampLoi,SynDampLoi)
t.test(AxeWetLoi,SynWetLoi)


```
