---
title: "An investigation of the long term ecological and soil carbon storage responses of the Upper Axe valley to landscape scale anthropogenic impacts"
author: "Julian Payne"
date: "2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
---


### MSc Land and Ecological Restoration - Research Project
# Variation of LOI with soil horizon, and by floodplain/paleochannel core location
### Wetlands split into BuriedWetlands (including RiverBed) located underneath grey clays and Wetlands for contemporary overlying wetland soils and associated grey clay

```{r}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```

### Load packages
```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(readxl)
library(cowplot)
library(vegan)
```

### Load the data for the project
```{r}
filepath<-here("data","Soil_Cores.xlsx")
soil_cores<-read_excel(filepath,
                       sheet = "SoildataV3",
                       range = "A1:Z155")
```

### Set soil horizon layer order
```{r}
HorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Wetland", "BuriedWetland")
PaleoHorizonOrder<-c("Floodplain1", "Gley", "Wetland", "Buried Wetland")
FldPlnHorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Buried Wetland")
WetnessOrder<-c("Dry", "Damp", "Wet")
```

### Clean up the data
Notice we have some missing values in the loi column, identified by 0 in test_for_loi

```{r}
soilloi_clean<-soil_cores %>%
  filter(horizon_type == "Floodplain2" | horizon_type == "Floodplain1" | horizon_type == "Gley" | horizon_type == "Wetland" | horizon_type == "BuriedWetland") %>% 
  select(-loi) %>%
  drop_na()
```

## split floodplain and paleochannel data
```{r}
paleochannel<-soilloi_clean %>%
  filter(core_type=="Paleochannel")
floodplain<-soilloi_clean %>%
  filter(core_type=="Floodplain")
```

# Summarise the data

```{r}
horizon_summary<-soilloi_clean %>%
  group_by(horizon_type) %>%
  summarise(mean.Sloi = mean(loi_calc),
            se.Sloi = sd(loi_calc/sqrt(n())))%>%
  arrange(HorizonOrder)
horizon_summary
```
```{r}
PaleoHorizonSummary<-paleochannel%>%
  group_by(horizon_type) %>%
  summarise(mean.Sloi = mean(loi_calc),
            se.Sloi = sd(loi_calc/sqrt(n())))%>%
  arrange(PaleoHorizonOrder)
PaleoHorizonSummary
```

```{r}
FloodHorizonSummary<-floodplain%>%
  group_by(horizon_type) %>%
  summarise(mean.Sloi = mean(loi_calc),
            se.Sloi = sd(loi_calc/sqrt(n())))%>%
  arrange(FldPlnHorizonOrder)
FloodHorizonSummary
```


# Plot the full dataset

```{r}
soilloi_clean%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=loi_calc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "Total Carbon LOI",
       title = "Variation of LOI by soil horizon") +
  theme_cowplot() 
```

## Test if soil horizon data sets are significantly different

```{r}
Fdpn1Loi<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  select(loi_calc) %>%
  pull() 

Fdpn2Loi<-soilloi_clean %>%
  filter(horizon_type=="Floodplain2") %>% 
  select(loi_calc) %>%
  pull() 

GleyLoi<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  select(loi_calc) %>%
  pull() 

WetlandLoi<-soilloi_clean %>%
  filter(horizon_type=="Wetland") %>%
  select(loi_calc) %>%
  pull()

BuriedWetLoi<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  select(loi_calc) %>%
  pull()

t.test(Fdpn1Loi,BuriedWetLoi)
t.test(Fdpn2Loi,BuriedWetLoi)
t.test(GleyLoi,BuriedWetLoi)
t.test(WetlandLoi,BuriedWetLoi)

t.test(Fdpn1Loi,Fdpn2Loi)
t.test(Fdpn1Loi,GleyLoi)
t.test(Fdpn1Loi, WetlandLoi)

t.test(GleyLoi, WetlandLoi)
t.test(GleyLoi, Fdpn2Loi)

t.test(WetlandLoi, Fdpn2Loi)

```

## Plot the LOI data by soil horizon and by core location (Floodplain/Paleochannel)

```{r}
soilloi_clean%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=loi_calc, ymax =0.35, ymin = 0, colour = core_type)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil Horizon",
       y = "Total Carbon LOI",
       title = "Variation of LOI with soil horizon and core location",
       colour = "Core location") +
  theme_cowplot()+
  theme(legend.position = c(0.05,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10)) 
```


## Test if floodplain/ paleochannel soil horizon data sets are significantly different

```{r}
Fdpn1FloodLoi<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>%
  filter(core_type=="Floodplain") %>%  
  select(loi_calc) %>%
  pull() 

Fdpn1PaleoLoi<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(core_type=="Paleochannel") %>%  
  select(loi_calc) %>%
  pull() 

GleyFloodLoi<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(core_type=="Floodplain") %>%  
  select(loi_calc) %>%
  pull() 

GleyPaleoLoi<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(core_type=="Paleochannel") %>%  
  select(loi_calc) %>%
  pull() 

BuriedWetFloodLoi<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(core_type=="Floodplain") %>%  
  select(loi_calc) %>%
  pull()

BuriedWetPaleoLoi<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(core_type=="Paleochannel") %>%  
  select(loi_calc) %>%
  pull()

t.test(Fdpn1FloodLoi, Fdpn1PaleoLoi)
t.test(GleyFloodLoi, GleyPaleoLoi)
t.test(BuriedWetFloodLoi, BuriedWetPaleoLoi)

```

