---
title: "An investigation of the long term ecological and soil carbon storage responses of the Upper Axe valley to landscape scale anthropogenic impacts"
author: "Julian Payne"
date: "2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### MSc Land and Ecological Restoration - Research Project
# Variation of LOI with core distance from the river
### Wetlands split into BuriedWetlands (including RiverBed) located underneath grey clays and Wetlands for contemporary overlying wetland soils and associated grey clay


```{r}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```

## Load packages
```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(readxl)
library(cowplot)
library(vegan)
library(lme4)
library(lmerTest)
```

## Load the data for the project
```{r}
filepath<-here("data","Soil_Cores.xlsx")
soil_cores<-read_excel(filepath,
                       sheet = "SoildataV3",
                       range = "A1:AB155")
```

## Set soil horizon layer order
```{r}
HorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Wetland", "BuriedWetland")
PaleoHorizonOrder<-c("Floodplain1", "Gley", "Wetland", "Buried Wetland")
FldPlnHorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Buried Wetland")
WetnessOrder<-c("Dry", "Damp", "Wet")
```

### Clean up the data
Notice we have some missing values in the loi column, identified by 0 in test_for_loi

```{r}
soilloi_clean<-soil_cores %>%
  filter(horizon_type == "Floodplain2" | horizon_type == "Floodplain1" | horizon_type == "Gley" | horizon_type == "Wetland" | horizon_type == "BuriedWetland") %>% 
  select(-loi) %>%
  drop_na()
```

## Mixed effects for horizon type, wetness and distance of core from the river

```{r}
lmSoil<-lmer(loi_calc~horizon_type * wetdrydamp + (horizon_type|DistFromRiver), data=soilloi_clean)
summary(lmSoil)
```

```{r}
  anova(lmSoil)
```

# For all samples

```{r}
soilloi_clean%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Total Carbon LOI",
       title = "Influence of distance from the river on all soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
  ylim(0,27) +
  xlim(5,66)+
  theme_cowplot() +
  theme(title = element_text(size=11.5))
```

```{r}
   DistFromRiverModel<-lm(log(loi_calc)~DistFromRiver, data=soilloi_clean)
   autoplot(DistFromRiverModel)[1:2] + theme_cowplot()
   anova(DistFromRiverModel)
```

# for gley soils

```{r}
soilloi_clean%>%
  filter(horizon_type=="Gley")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Total Carbon LOI",
       title = "Influence of distance from the river on gley soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
  ylim(0,27) +
  xlim(5,66)+
  theme_cowplot()  +
  theme(axis.title.y = element_text(hjust=0.05))+
  theme(title = element_text(size=11.5))
```

# Linear fit for Gley soils

```{r}
GleyfromRiverSoil<-soilloi_clean%>%
   filter(horizon_type=="Gley")  
   GleyfromRiverModel<-lm(loi_calc~DistFromRiver, data=GleyfromRiverSoil)
   autoplot(GleyfromRiverModel)[1:2] + theme_cowplot()
   anova(GleyfromRiverModel)
```

# Floodplain1 soils

```{r}
soilloi_clean%>%
  filter(horizon_type=="Floodplain1")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Total Carbon LOI",
       title = "Influence of distance from the river on Floodplain1 soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
  ylim(0,27) +
  xlim(5,66)+
  theme_cowplot()  +
  theme(axis.title.y = element_text(hjust=0.05))+
  theme(title = element_text(size=11.5))
```

# linear fit for Floodplain1 soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(horizon_type=="Floodplain1")  
   WetfromRiverModel<-lm(loi_calc~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```


```{r}
soilloi_clean%>%
  filter(wetdrydamp=="Wet")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Carbon LOI",
       title = "Influence of distance from the river on wet soil samples") +
#  xlim(5,40)+
  xlim(5,66)+
  ylim(0,27) +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot()  +
  theme(title = element_text(size=11.5))
```

# log fit for wet soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(wetdrydamp=="Wet")  
   WetfromRiverModel<-lm(log(loi_calc)~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```

```{r}
soilloi_clean%>%
  filter(horizon_type=="BuriedWetland")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Carbon LOI",
       title = "Influence of distance from the river on Buried Wetland soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
#  ylim(0,10)+
  ylim(0,27) +
  xlim(5,66)+
  theme_cowplot()  +
  theme(title = element_text(size=11.5))
```

# log fit for buried wetland soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(horizon_type=="BuriedWetland")  
   WetfromRiverModel<-lm(log(loi_calc)~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```


```{r}
soilloi_clean%>%
  filter(horizon_type=="BuriedWetland")%>%
  filter(wetdrydamp=="Wet")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Carbon LOI",
       title = "Influence of distance from the river on wet Buried Wetland soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
#  ylim(0,10)+
  ylim(0,27) +
  xlim(5,66)+
  theme_cowplot()  +
  theme(title = element_text(size=11.5))
```

# log fit for wet buried wetland soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(wetdrydamp=="Wet")%>%
   filter(horizon_type=="BuriedWetland")  
   WetfromRiverModel<-lm(log(loi_calc)~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```

```{r}
soilloi_clean%>%
  filter(horizon_type=="Wetland")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Carbon LOI",
       title = "Influence of distance from the river on Wetland soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
#  ylim(0,10)+
  ylim(0,27) +
  xlim(5,66)+
  theme_cowplot()  +
  theme(title = element_text(size=11.5))
```

# log fit for wetland soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(horizon_type=="Wetland")  
   WetfromRiverModel<-lm(loi_calc~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```

# Floodplain2 soils

```{r}
soilloi_clean%>%
  filter(horizon_type=="Floodplain2")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Carbon LOI",
       title = "Influence of distance from the river on Floodplain2 soil samples") +
  geom_smooth(method = "lm", se=FALSE)+
  ylim(0,27) +
#  ylim(0,10)+
#  xlim(5,40)+
  xlim(5,66)+
  theme_cowplot()  +
  theme(title = element_text(size=11.5))
```

# log fit for Floodplain2 soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(horizon_type=="Floodplain2")  
   WetfromRiverModel<-lm(loi_calc~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```

```{r}
soilloi_clean%>%
  filter(horizon_type=="Gley")%>%
  filter(wetdrydamp=="Wet")%>%
  ggplot(aes(x=DistFromRiver, y=loi_calc)) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance from the river (m)",
       y = "Carbon LOI",
       title = "Influence of distance from the river on wet gley soil samples") +
  ylim(0,27) +
#  ylim(0,10)+
  xlim(5,66)+
#  xlim(5,40)+
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot()  +
  theme(title = element_text(size=11.5))
```

# log fit for wet Gley soils

```{r}
WetfromRiverSoil<-soilloi_clean%>%
   filter(horizon_type=="Gley")%>%
   filter(wetdrydamp=="Wet")  
   WetfromRiverModel<-lm(loi_calc~DistFromRiver, data=WetfromRiverSoil)
   autoplot(WetfromRiverModel)[1:2] + theme_cowplot()
   anova(WetfromRiverModel)
```

