---
title: "An investigation of the long term ecological and soil carbon storage responses of the Upper Axe valley to landscape scale anthropogenic impacts"
author: "Julian Payne"
date: "2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### MSc Land and Ecological Restoration - Research Project
# Variation of LOI along the Synderford valley, by depth, soil horizon, and wetness
### Wetlands split into BuriedWetlands (including RiverBed) located underneath grey clays and Wetlands for contemporary overlying wetland soils and associated grey clay

```{r}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```

## Load packages
```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(readxl)
library(cowplot)
library(vegan)
library(lme4)
library(lmerTest)
```

## Load the data for the project
```{r}
filepath<-here("data","Soil_Cores.xlsx")
soil_cores<-read_excel(filepath,
                       sheet = "SoildataV3",
                       range = "A1:AB155")
```

## Set soil horizon layer order
```{r}
HorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Wetland", "BuriedWetland")
PaleoHorizonOrder<-c("Floodplain1", "Gley", "Wetland", "Buried Wetland")
FldPlnHorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Buried Wetland")
WetnessOrder<-c("Dry", "Damp", "Wet")
```

### Clean up the data
Notice we have some missing values in the loi column, identified by 0 in test_for_loi

```{r}
soilloi_clean<-soil_cores %>%
  filter(horizon_type == "Floodplain2" | horizon_type == "Floodplain1" | horizon_type == "Gley" | horizon_type == "Wetland" | horizon_type == "BuriedWetland") %>% 
  select(-loi) %>%
  drop_na()
```

## Plot the data by soil horizon

```{r}
soilloi_clean%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=loi_calc, colour = factor( horizon_type, ordered(HorizonOrder)))) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Carbon LOI",
       title = "Variation of LOI for along the length of the Synderford Valley",
       colour = "Soil horizon") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=11.5))
```


```{r}
soilloi_clean%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=loi_calc, colour = factor( wetdrydamp, ordered(WetnessOrder)))) +
  geom_point(size=2, alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Carbon LOI",
       title = "Variation of LOI for along the length of the Synderford Valley",
       colour = "Soil wetness") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=11.5))
```


# Mixed effects for horizon type, soil wetness and distance along the the Synderford Valley are not possible dur to limited sample numbers

# Error: number of observations (=31) <= number of random effects (=32) for term (horizon_type | Distance); the random-effects parameters and the residual variance (or scale parameter) are probably unidentifiable

# Graphical representation of the data by Horizon type and by wetness

```{r}
soilloi_clean%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=-depth_cm, colour = factor( horizon_type, ordered(HorizonOrder)))) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of LOI with Depth and distance along the Synderford Valley",
       size ="Loi",
       colour = "Soil horizon") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```


```{r}
soilloi_clean%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=-depth_cm, colour = factor( wetdrydamp, ordered(WetnessOrder)))) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of LOI with Depth and distance along the Synderford Valley",
       size ="Loi",
       colour = "Soil wetness") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```

# Investigation of each 'significant' relationship

# wetness - dry, damp, wet

```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(wetdrydamp=="Dry")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```


```{r}
soilloi_clean%>%
  filter(wetdrydamp=="Dry")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=loi_calc )) +
  geom_point(size = 3, alpha=0.5, colour = "Red")+
    labs(x = "Distance along the Axe valley (m from downstream)",
       y = "Carbon LOI",
       title = "Variation of dry soils' LOI with distance along the Synderford Valley") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```


```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(wetdrydamp=="Damp")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```

```{r}
soilloi_clean%>%
  filter(wetdrydamp=="Damp")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=loi_calc )) +
  geom_point(size = 3, alpha=0.5, colour = "Green")+
    labs(x = "Distance along the Axe valley (m from downstream)",
       y = "Carbon LOI",
       title = "Variation of damp soils' LOI with distance along the Synderford Valley") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```

```{r}
soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(wetdrydamp=="Wet")%>%
  ggplot(aes(x=Distance, y=-depth_cm, colour = factor( horizon_type, ordered(HorizonOrder)))) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of wet soil's LOI with Depth and distance along the Synderford Valley",
       size ="Loi",
       colour = "Soil horizon") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```

```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(wetdrydamp=="Wet")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```

# for gley soils

```{r}
soilloi_clean%>%
  filter(horizon_type=="Gley")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=loi_calc )) +
  geom_point(size = 3, alpha=0.5, colour = "Green")+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Carbon LOI",
       title = "Variation of Gley soils' LOI with distance along the Synderford Valley") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```

```{r}
soilloi_clean%>%
  filter(horizon_type=="Gley")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=-depth_cm )) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of Gley soils' LOI with Depth and distance along the Synderford Valley",
       size ="Loi") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10))
```


```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(horizon_type=="Gley")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```


```{r}
soilloi_clean%>%
  filter(horizon_type=="Gley")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=-depth_cm, colour = factor( wetdrydamp, ordered(WetnessOrder)))) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of Gley soils' LOI with Depth and distance along the Synderford Valley",
       size ="Loi",
       colour = "Soil wetness") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10))
```


```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(horizon_type=="Gley")%>%
  filter(wetdrydamp=="Dry")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```


```{r}
soilloi_clean%>%
  filter(horizon_type=="Wetland")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=-depth_cm, colour = core_type)) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of Wetland soils' LOI with Depth and distance along the Synderford Valley",
       size ="Loi",
       colour = "Soil horizon") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=10.5))
```

```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(horizon_type=="Wetland")
```


```{r}
soilloi_clean%>%
  filter(horizon_type=="BuriedWetland")%>%
  filter(River=="Synderford")%>%
  ggplot(aes(x=Distance, y=-depth_cm, colour = core_type)) +
  geom_point(aes(size=loi_calc), alpha=0.5)+
    labs(x = "Distance along the Synderford valley (m from downstream)",
       y = "Depth below surface",
       title = "Variation of Buried Wetland soils' LOI with Depth and distance along the Synderford Valley",
       size ="Loi",
       colour = "Soil horizon") +
  geom_smooth(method = "lm", se=FALSE)+
  theme_cowplot() +
  theme(title = element_text(size=9))
```


```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(horizon_type=="BuriedWetland")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```


```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(horizon_type=="BuriedWetland")%>%
  filter(core_type=="Floodplain")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```


```{r}
DepthDist<-soilloi_clean%>%
  filter(River=="Synderford")%>%
  filter(horizon_type=="BuriedWetland")%>%
  filter(core_type=="Paleochannel")
```

```{r}
Distmodel<-lm(log(loi_calc)~Distance,data=DepthDist)
autoplot(Distmodel)[1:2] + theme_cowplot()
anova(Distmodel)
```

