---
title: "An investigation of the long term ecological and soil carbon storage responses of the Upper Axe valley to landscape scale anthropogenic impacts"
author: "Julian Payne"
date: "2024"
output:
  html_document:
    df_print: paged
  pdf_document: default
---

### MSc Land and Ecological Restoration - Research Project
# Variation of NPOC by week, soil horizon and soil wetness
## For Shimadzu, data only available for weeks 1 and 3
## Wetlands split into BuriedWetlands (including RiverBed) located underneath grey clays
## and Wetlands for contemporary overlying wetland soils and associated grey clay


```{r}
knitr::opts_chunk$set(message=FALSE,warning=FALSE)
```

## Load packages
```{r}
library(tidyverse)
library(here)
library(ggfortify)
library(readxl)
library(cowplot)
library(vegan)
```

## Load the data for the project
```{r}
filepath<-here("data","Soil_Cores.xlsx")
soil_cores<-read_excel(filepath,
                       sheet = "SoildataV3",
                       range = "A1:AE155")
```

## Set soil horizon layer order
```{r}
HorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "Wetland", "BuriedWetland")
PaleoHorizonOrder<-c("Floodplain1", "Gley", "Wetland", "BuriedWetland")
FldPlnHorizonOrder<-c("Floodplain1", "Floodplain2", "Gley", "BuriedWetland")
WetnessOrder<-c("Dry", "Damp", "Wet")
```

### Clean up the data
Notice we have some missing values in the loi column, identified by 0 in test_for_loi

```{r}
soilloi_clean<-soil_cores %>%
  filter(horizon_type == "Floodplain2" | horizon_type == "Floodplain1" | horizon_type == "Gley" | horizon_type == "Wetland" | horizon_type == "BuriedWetland") %>% 
  select(-fresh_weight) %>%
  drop_na()
```

## Summarise the data

```{r}
horizon_summary<-soilloi_clean %>%
  group_by(horizon_type) %>%
  summarise(mean.Snpoc = mean(npoc),
            se.Snpoc = sd(npoc/sqrt(n())))%>%
  arrange(HorizonOrder)
horizon_summary
```


```{r}
horizon_summary<-soilloi_clean %>%
  group_by(horizon_type) %>%
  filter(Week=="Week1")%>%
  summarise(mean.Snpoc = mean(npoc),
            se.Snpoc = sd(npoc/sqrt(n())))%>%
  arrange(HorizonOrder)
horizon_summary
```

```{r}
horizon_summary<-soilloi_clean %>%
  group_by(horizon_type) %>%
  filter(Week=="Week3")%>%
  summarise(mean.Snpoc = mean(npoc),
            se.Snpoc = sd(npoc/sqrt(n())))
horizon_summary
```


# Plot the data by soil horizon

```{r}
soilloi_clean%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0, colour=Week)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil Horizon",
       y = "NPOC",
       title = "Variation of NPOC by soil horizon") +
  ylim(0,115) +
  theme_cowplot() +
  theme(legend.position = c(0.01,0.70), legend.title = element_text(size =12), legend.text = element_text(size=10))
```


```{r}
soilloi_clean%>%
  filter(Week=="Week1")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20", alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "NPOC",
       title = "Variation of Week 1 NPOC by soil horizon") +
  ylim(0,115) +
  theme_cowplot() 
```


## Test if soil horizon data sets are significantly different in Week 1

```{r}
Fdpn1npoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

Fdpn2npoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain2") %>% 
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

Gleynpoc<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

Wetlandnpoc<-soilloi_clean %>%
  filter(horizon_type=="Wetland") %>%
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull()

BuriedWetnpoc<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull()

t.test(Fdpn1npoc,BuriedWetnpoc)
t.test(Fdpn2npoc,BuriedWetnpoc)
t.test(Gleynpoc,BuriedWetnpoc)
t.test(Wetlandnpoc,BuriedWetnpoc)

t.test(Fdpn1npoc,Fdpn2npoc)
t.test(Fdpn1npoc,Gleynpoc)
t.test(Gleynpoc, Wetlandnpoc)

```


```{r}
soilloi_clean%>%
  filter(Week=="Week3")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "NPOC",
       title = "Variation of Week 3 NPOC by soil horizon") +
  ylim(0,115) +
  theme_cowplot() 
```

## Test if soil horizon data sets are significantly different in Week 3

```{r}
Fdpn1npoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

Fdpn2npoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain2") %>% 
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

Gleynpoc<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

Wetlandnpoc<-soilloi_clean %>%
  filter(horizon_type=="Wetland") %>%
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull()

BuriedWetnpoc<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull()

t.test(Fdpn1npoc,BuriedWetnpoc)
# t.test(Fdpn2npoc,BuriedWetnpoc)
t.test(Gleynpoc,BuriedWetnpoc)
# t.test(Wetlandnpoc,BuriedWetnpoc)

# t.test(Fdpn1npoc,Fdpn2npoc)
t.test(Fdpn1npoc,Gleynpoc)
# t.test(Gleynpoc, Wetlandnpoc)

```

# Plot the data by soil wetness

```{r}
soilloi_clean%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=npoc, ymax =0.35, ymin = 0, colour = Week)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil wetness",
       y = "NPOC",
       title = "Variation of NPOC by soil Wetness") +
  ylim(0,115) +
  theme_cowplot() +
  theme(legend.position = c(0.01,0.85), legend.title = element_text(size =12), legend.text = element_text(size=10))
```

## For week 1

```{r}
soilloi_clean%>%
  filter(Week=="Week1")%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=npoc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil wetness",
       y = "NPOC",
       title = "Variation of Week 1 NPOC by soil Wetness") +
  ylim(0,115) +
  theme_cowplot() 
```

## Test if soil wetness data sets are significantly different in Week 1

```{r}
Wetnpoc<-soilloi_clean %>%
  filter(wetdrydamp=="Wet") %>% 
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

Dampnpoc<-soilloi_clean %>%
  filter(wetdrydamp=="Damp") %>% 
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

Drynpoc<-soilloi_clean %>%
  filter(wetdrydamp=="Dry") %>% 
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

t.test(Wetnpoc, Dampnpoc)
t.test(Wetnpoc, Drynpoc)
t.test(Dampnpoc, Drynpoc)
```

## week 3

```{r}
soilloi_clean%>%
  filter(Week=="Week3")%>%
  ggplot(aes(x=factor(wetdrydamp,WetnessOrder), y=npoc, ymax =0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil wetness",
       y = "NPOC",
       title = "Variation of Week 1 NPOC by soil Wetness") +
  ylim(0,115) +
  theme_cowplot() 
```

## Test if soil wetness data sets are significantly different in Week 3

```{r}
Wetnpoc<-soilloi_clean %>%
  filter(wetdrydamp=="Wet") %>% 
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

Dampnpoc<-soilloi_clean %>%
  filter(wetdrydamp=="Damp") %>% 
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

Drynpoc<-soilloi_clean %>%
  filter(wetdrydamp=="Dry") %>% 
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

t.test(Wetnpoc, Dampnpoc)
t.test(Wetnpoc, Drynpoc)
t.test(Dampnpoc, Drynpoc)

```


## For Week 1 for horizon and river
 
```{r}
soilloi_clean%>%
  filter(Week=="Week1")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0, colour = River)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Variation of Week 1 npoc with soil horizon, and river") +
  theme_cowplot()+
  ylim(0,115) +
  theme(legend.position = c(0.1,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10)) 
```

```{r}
soilloi_clean%>%
  filter(Week=="Week1")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0, colour = core_type)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Variation of Week1 npoc with soil horizon and core location",
       colour = "Core location") +
  theme_cowplot()+
  ylim(0,115) +
  theme(legend.position = c(0.1,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10)) 
```


```{r}
soilloi_clean%>%
  filter(Week=="Week1")%>%
  filter(core_type=="Paleochannel")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax = 0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Week 1 npoc for Paleochannel core soil samples") +
  ylim(0,115) +
  theme_cowplot() 
```

```{r}
soilloi_clean%>%
  filter(Week=="Week1")%>%
  filter(core_type=="Floodplain")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax = 0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Week 1 npoc for Floodplain core soil samples") +
  ylim(0,115) +
  theme_cowplot() 
```

## Test if Week 1 floodplain/ paleochannel soil horizon data sets are significantly different

```{r}
Fdpn1Floodnpoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>%
  filter(core_type=="Floodplain") %>%  
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

Fdpn1Paleonpoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(core_type=="Paleochannel") %>%  
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

GleyFloodnpoc<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(core_type=="Floodplain") %>%  
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

GleyPaleonpoc<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(core_type=="Paleochannel") %>%  
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull() 

BuriedWetFloodnpoc<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(core_type=="Floodplain") %>%  
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull()

BuriedWetPaleonpoc<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(core_type=="Paleochannel") %>%  
  filter(Week=="Week1")%>%
  select(npoc) %>%
  pull()

t.test(Fdpn1Floodnpoc, Fdpn1Paleonpoc)
t.test(GleyFloodnpoc, GleyPaleonpoc)
t.test(BuriedWetFloodnpoc, BuriedWetPaleonpoc)

```

## For Week 3
 
```{r}
soilloi_clean%>%
  filter(Week=="Week3")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0, colour = River)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Variation of Week 3 npoc with soil horizon, and river") +
  theme_cowplot()+
  ylim(0,115) +
  theme(legend.position = c(0.8,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10)) 
```

```{r}
soilloi_clean%>%
  filter(Week=="Week3")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax =0.35, ymin = 0, colour = core_type)) +
  geom_boxplot()+
  geom_point(position = position_jitterdodge())+
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Variation of Week 3 npoc with soil horizon and core location",
       colour = "Core location") +
  theme_cowplot()+
  ylim(0,115) +
  theme(legend.position = c(0.8,0.9), legend.title = element_text(size =12), legend.text = element_text(size=10)) 
```

```{r}
soilloi_clean%>%
  filter(Week=="Week3")%>%
  filter(core_type=="Paleochannel")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax = 0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Week 3 npoc for Paleochannel core soil samples") +
  ylim(0,115) +
  theme_cowplot() 
```

```{r}
soilloi_clean%>%
  filter(Week=="Week3")%>%
  filter(core_type=="Floodplain")%>%
  ggplot(aes(x=reorder(horizon_type,Horizon_order), y=npoc, ymax = 0.35, ymin = 0)) +
  geom_boxplot()+
  geom_jitter(width=0.1, colour = "#f03b20",alpha=0.5) +
  labs(x = "Soil Horizon",
       y = "npoc",
       title = "Week 3 npoc for Floodplain core soil samples") +
  ylim(0,115) +
  theme_cowplot() 
```

## Test if Week 3 floodplain/ paleochannel soil horizon data sets are significantly different

```{r}
Fdpn1Floodnpoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>%
  filter(core_type=="Floodplain") %>%  
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

Fdpn1Paleonpoc<-soilloi_clean %>%
  filter(horizon_type=="Floodplain1") %>% 
  filter(core_type=="Paleochannel") %>%  
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

GleyFloodnpoc<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(core_type=="Floodplain") %>%  
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

GleyPaleonpoc<-soilloi_clean %>%
  filter(horizon_type=="Gley") %>% 
  filter(core_type=="Paleochannel") %>%  
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull() 

BuriedWetFloodnpoc<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(core_type=="Floodplain") %>%  
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull()

BuriedWetPaleonpoc<-soilloi_clean %>%
  filter(horizon_type=="BuriedWetland") %>%
  filter(core_type=="Paleochannel") %>%  
  filter(Week=="Week3")%>%
  select(npoc) %>%
  pull()

# t.test(Fdpn1Floodnpoc, Fdpn1Paleonpoc)
t.test(GleyFloodnpoc, GleyPaleonpoc)
t.test(BuriedWetFloodnpoc, BuriedWetPaleonpoc)

```
